Tabular Constraint Learning

### 2016-04-14 Paper/ECAI/tabular.tex
added the discussion section with comments
Small corrections

\tias{Rewrite}
\section{Introduction}
Millions of people across use spreadsheets everyday. Yet, many people lack a complete understanding of the structures and existing dependencies in the data. Indeed, the most useful part of spreadsheets is often the formulas and often these formula disappear, when a file is exported and shared as a CSV file. This brings us to the question: can we discover or reconstruct structural relations in flat tabular spreadsheet data? [in a general way that allows declarative specification of constraints to discover]

Let us look at an example in Figure \ref{fig:main_example}, the name already suggest the usage of standard spreadsheet operations such as \textit{average} or \textit{sum}. If we look at the Table 1 and at the first row in the Table 2, it is clear that one could have selected the column called \textbf{1st Quarter}, clicked on the \textit{SUM} function and then just used the drag fill handle to apply the formula to the other columns. Similarly, that would work for the second (\textit{Average}), third (\textit{Min})  and forth (\textit{Max}) rows in Table 2. I.e., all operations are vectorized and make use of the standard spreadsheet operations.

It is not hard to see that computation goes not only in columns but in rows as well: if we sum the values in Table 1 in the first row in all four quarters, we would obtain the seventh column in Table 1. That gives us already a hint how this problem is different from the standard data mining settings, when the data is in rows and variables are in columns. Here everything is mixed, some values are missing in the results, some missing the arguments, even a header might have missing values. The data is relational on one hand, since we have multiple table with relationships between them. And on the other hand, the data is numeric, since people tend to use spreadsheets for financial and accounting computations, which is heavily numeric.

This makes tabular data understanding and constraint learning new and challenging, and it is not only challenging for a machine but for a human as well. Due to the complexity of numeric computations in spreadsheets, people often fail to grasp the properties of the data, which leads to so called the \textit{Spreadsheet risk}\footnote{\url{https://en.wikipedia.org/wiki/Spreadsheet\#Spreadsheet_risk}} and even caused major flaws in the famous economics papers \cite{flaw_excel} and billion-losses in the financial industry \cite{spreadsheet_risk_loss}. Tabular constraint learning provides a potential cure by indicating learned constraint violations and suggest possible fixes.

@@

\samuel{Spreadsheet risks: financial losses due to errors in spreadsheets}

@@

\section{Formalization}
\samuel{Explain concepts on example}
\sergey{subgroup notation, rework the solution definition}
\subsection{Groups, Type-consistency and Constraints}
In this work we focus on the spreadsheet data, which is physically one big table, but conceptually consists of multiple tables, see Figure \ref{fig:main_example} for an example. Formally, a table is $n \times m$ matrix, each entry is called a \textit{cell}. A cell has a type, e.g. namely numeric. In this work we distinguish the two types of data: numeric and textual. The numeric type has two subtypes: integers and floats. We also consider the special element called \textit{None}, which has two types: numeric and textual.

Based on these types we introduce a notation of type consistency for a set of elements. A set is called \textit{type-consistent} iff all elements are either numeric or textual. Certain constraints, such as \textit{rank} or \textit{series}, make use of the numeric sub-types by requiring its arguments to be integers.

We use the notation $T[a,b]$ to refer to a particular cell in the table $T$ at the row with index $a$ and the column with index $b$. To refer to the $a$-th row (column) we use the following syntax $T[a,{:}]$ ($T[{:},a]$), to refer to a subrange from $a$ to $b$ we use $T[a{:}b,]$ ($T[,a{:}b]$).

A \textit{vector} is either a column or a row that is type-consistent. If a vector is a row (column), we say that it has a \textit{row} (\textit{column}) orientation. A \textit{group} is a subrange of vectors with the same orientation in a table. We use the following notation to refer to a row (column) group $G$ in a table $T$ with rows (columns) ranging from $a$ to $b$: $G = T[a{:}b,:]$ ($G = T[{:},a{:}b]$), where $a,b$ are natural numbers. We denote as the \textit{length} of a row (column) group $G$, written as \textit{length(G)}, the number of its columns (rows). We call a group $G$ \textit{numeric} (\textit{textual}, etc), written as \textit{numeric(G)}, if its vectors contain numeric (textual, etc) elements.

We refer to a subgroup of a group $G$ in \groups as $g$. A \textit{subgroup} $g$ is a subrange of vectors in a group $G$. If a subgroup $g$ must constain only one vector, we write $g \in G$.

A \textit{Spreadsheet constraint} is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. \textit{\CName} is the textual name of the constraint together with the list of its variables $v_1,\dots,v_n$. \textit{\CSignature} is a constraint specifying the properties of the group assignments, corresponding to the variables $v_1,\dots,v_n$ in \CName, such as their types, e.g., requiring them to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are constraints on the group meta-information not on the actual group content. \textit{\CFunction} is a constraint specifying that the data in the subgroups satisfies the function. For example, a spreadsheet constraint \textit{rank} has \CName \textit{$v_y$ = RANK($v_x$)}, where $v_x$ and $v_y$ are the variables; its \CSignature is: the subgroup assigned to $v_x$ must be numeric, $v_y$ must be integer and the length of vectors in $v_x$ is the same as in $v_y$; its \CFunction is the following constraint: the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ is a solution iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_x$ has the rank (possibly with ties) specified in $g_y$.

Let us now define what is a tabular solution for a spreadsheet constraint in general. Let $C$ denote a spreadsheet constraint, $S$ a mapping from variables to the subgroups of \groups, and $C^S$ the constraint $C$, in which all variables are substituted with the corresponding subgroups from $S$, then $S$ is a \textit{tabular solution} to $C$ iff both $\CSignature^S$ and $\CFunction^S$ hold.

@@

\paragraph{Example}
Consider the \ecfkey{FK}{PK} (foreign-key) constraint: for every solution the subgroup (vector) $v_{PK}$ assigned to \textit{PK} must fulfill \ecalldiff{v_{PK}}.
Therefore, we say indicate \textit{foreign-key} as being dependent on \textit{all-different} and the latter will be treated first.
The \ecrank{v_y}{v_x} constraint does not have any dependencies, however in a restricted version without ties the vector $v_y$ would always be a permutation of numbers $1, 2, ..., \textit{length}(v_y)$.

@@

\subsection{Case study}
Let us illustrate our implementation using the example presented in Figure~\ref{fig:main_example}.
This example combines several smaller examples that were used in an exercise session to teach Excel into one spreadsheet.
Figure~\ref{fig:sol_example} shows the constraints that we expect to find.

\begin{figure}
  \label{fig:sol_example}
  {\small
    \begin{align*}
      & SERIES(\range{T_{1}}{\rangeall}{1}) \\
%
      & \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      & \ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecaggc{AVERAGE}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MAX}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MIN}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggif{SUM}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \ecaggif{MAX}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      & \range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}
  }
  \caption{Expected constraints in the case study \sergey{in Figure \ref{fig:main_example}; some more explanation needed, not self explanatory}}
\end{figure}

>>>>

\tias{Spreadsheet is an instance of tabular data, but not the only one?}

\tias{Rewrite}
\section{Introduction}
Millions of people across the world use spreadsheets everyday.
Yet, many people lack a complete understanding of the structures and existing dependencies in the data.
Indeed, the most useful part of spreadsheets is often the formulas and often these formula disappear when a file is exported and shared, for example as a CSV file.
This brings us to the question: can we discover or reconstruct structural relations in flat tabular spreadsheet data?
Moreover, can we accomplish this task in a general way that allows declarative specification of constraints to discover?

\samuel{Maybe slightly more formal:}
Let us look at an example in Figure~\ref{fig:main_example}, the header names already suggest the usage of standard spreadsheet operations such as \textit{average} or \textit{sum}.
If we look at Table~1 and at the first row in Table~2, it is clear that one could have selected the column called \textbf{1st Quarter}, clicked on the \textit{SUM} function and then just used the drag fill handle to apply the formula to the other columns.
Similarly, that would work for the second (\textit{Average}), third (\textit{Min})  and forth (\textit{Max}) rows in Table 2.
I.e., all operations are vectorized and make use of the standard spreadsheet operations.

Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.
\samuel{Maybe slightly more formal:}
That gives us already a hint how this problem is different from the standard data mining settings, when the data is in rows and variables are in columns. Here everything is mixed, some values are missing in the results, some missing the arguments, even a header might have missing values. The data is relational on one hand, since we have multiple table with relationships between them. And on the other hand, the data is numeric, since people tend to use spreadsheets for financial and accounting computations, which is heavily numeric.

Given these novel aspects, we conclude that understanding and learning constraints in tabular data is a new and challenging problem setting.
Futhermore, it is not only challenging for a machine but for a human as well.
\samuel{Maybe slightly more formal:} Due to the complexity of numeric computations in spreadsheets, people often fail to grasp the properties of the data, which leads to so called the Spreadsheet risk\footnote{\url{https://en.wikipedia.org/wiki/Spreadsheet\#Spreadsheet_risk}} and even caused major flaw in the famous economics papers \cite{flaw_excel} and caused billion-losses in the financial industry. Tabular constraint learning provides a potential cure by indicating learned constraint violations and suggest possible fixes.

@@

@@

\section{Formalization}
\samuel{Explain concepts on example}
\sergey{subgroup notation, rework the solution definition}
\subsection{Groups, Type-consistency and Constraints}
In this work we focus on spreadsheet data, which is physically one big table, but conceptually often consists of multiple tables, see Figure~\ref{fig:main_example} for an example.
Formally, a table is an $n \times m$ matrix, each table entry is called  a \textit{cell}.

Each cell has a {\em type} which can be numeric or textual.
The numeric type has two subtypes: integers and floats. 
We also consider the special element called \textit{None}, which can be either type: numeric or textual.
A set of cells is called \textit{type-consistent} iff all cells in the set are either numeric or textual. 
%Certain constraints, such as \textit{rank} or \textit{series}, make use of the numeric subtype by requiring its arguments to be integers.

We consider various sets of cells in tables: 
\begin{itemize}
  \item {\bf Cells}: Each entry in the matrix/table is called  a \textit{cell}.
  We use the notation $T[a,b]$ to refer to a particular cell in the table~$T$ at the row with index~$a$ and the column with index~$b$. 
  \item {\bf Columns and rows}: To refer to the $a$-th row (column) we use the following syntax $T[a,{:}]$ ($T[{:},a]$), to refer to a subrange from $a$ to $b$ we use $T[a{:}b,]$ ($T[,a{:}b]$).
  \item 
  A \textit{vector} is either a column or a row that is type-consistent. 
  If a vector is a row (column), we say that it has a \textit{row} (\textit{column}) orientation.
  \item
  A \textit{group} is a subrange of vectors with the same orientation in a table.
  We use the following notation to refer to a row (column) group $G$ in a table $T$ with rows (columns) ranging from $a$ to $b$: $G = T[a{:}b,:]$ ($G = T[{:},a{:}b]$), where $a,b$ are natural numbers.
  We denote as the \textit{length} of a row (column) group $G$, written as \textit{length(G)}, the number of its columns (rows). We call a group $G$ \textit{numeric} (\textit{textual}, etc), written as \textit{numeric(G)}, if \luc{all?????} its vectors contain numeric (textual, etc) elements.
  \item
  A \textit{subgroup} $G'$ is a subrange of vectors belonging to a group $G$. If the subgroup $\{v\}$ must contain only one vector, we write $v \in G$.
\end{itemize}
For notational convenience, we will refer to cells with $C$,  to vectors with $V$, and to groups and subgroups with $G$. 
When we do not want to distinguish between cells, columns, rows, vectors or groups, we shall talk about a {\em set of cells}.
\luc{add an example of group/ subgroup}

\tias{Use generic constraint}
A \textit{constraint} definition is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. 
\begin{itemize}
\item
\textit{\CName} is the name of the constraint together with the name of the groups, 
\item \textit{\CSignature} is specifying a set of restrictions on the types and variables of the constraint, 
% the properties of the group assignments, corresponding to the variables in \CName, such as their types, 
e.g., requiring the variables to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are restrictions on the group meta-information not on the actual group content. 
\item \textit{\CFunction} defines the actual constraint that needs to be satisfied in the data.
\end{itemize}

\paragraph{Example}
The constraint \textit{rank} has 
\begin{itemize}
\item \CName = \textit{$g_Y$ = RANK($g_X$)}, where $g_X$ and $g_Y$ are the group variables; 
\item \CSignature = subgroup $g_X$ (associated with X) is numeric, $g_Y$ is integer and the length of vectors in $g_X$ is the same as in $g_Y$; and
\item
\CFunction = \textit{$g_Y$ = RANK($g_X$)} iff $g_X \in G_X, g_Y \in G_Y$ and each value in $g_X$ has the rank (possibly with ties) specified in $g_Y$.
\end{itemize}

\samuel{2nd problem statement (from approach), remove EQUAL}

We can now define a classical constraint satisfaction problem on tables. 
A {\em constraint satisfaction problem} (CSP)  consists of 1) a set of cells, and 2) a set of constraints over those cells. 
Notice that the constraints specify the types (or the domains) that one encounters in traditional CSPs.  
Notice also, that as in traditional CSPs, the CSP problem defines the variables (the cells) of the problem
as well as the constraint that should hold amongst them. The task of CSP-solvers is then to find an assignment 
to the variables that satisfies all the constraints. The distinction between variables (cells) and their assignments (their values)
is important here. So, when we speak about cells, vectors and groups, we refer to the variables;
when we talk about values, assignments or instances, we refer to the values these cells take in a particular "instantiated" table.

While the traditional CSP problem is given a set of cells and a set of constraints,
and the task is to find an assignment, the {\bf inverse CSP} or the {\bf Tabular Constraint Learning Problem}.
the can be defined as follows.\\
\\
{\bf Given }
\begin{itemize}
\item
a set of instantiated tables (so all cells have values) ${\cal T}$; 
\item
a set of possible constraints ${\cal C}$; 
\item
a set of possible groups for the tables ${\cal G}$;
\end{itemize}
\noindent
{\bf Find}  the set $C\subset {\cal C}$ that contains all constraints $c(G_1, ... , G_N)$ that 1) are satisfied  in ${\cal T}$, 2)
for which
all $G_i \in {\cal G}$ and  3)  $c(G_1, ... , G_N)$ satisfies the signature of the constraint $c$.

\luc{ADD/REFER TO EXAMPLE}

%A \textit{tabular solution} $S$ to a spreadsheet constraint $C$ (with $n$ variables) is a subgroup assignment $(g_1, \dots, g_n)$ where $g_i$, such that $(g_1,\dots,g_n)$ satisfies both the \CSignature and \CFunction of the constraint $C$. \sergey{to finish this definition later, need to say that $g_i$ is the subgroup assigned to the $i$-th variable in $C$}

@@

\paragraph{Example}
Consider the \ecfkey{FK}{PK} (foreign-key) constraint: for every solution the subgroup (vector) $G_{PK}$ assigned to $PK$ must fulfill \ecalldiff{G_{PK}}.
Therefore, we say indicate \textit{foreign-key} as being dependent on \textit{all-different} and the latter will be treated first.
The \ecrank{Y}{X} constraint does not have any dependencies, however in a restricted version without ties the vector $Y$ would always be a permutation of numbers $1, 2, ..., length(Y)$.

@@

\subsection{Case study}
Let us illustrate our implementation using the example presented in figure~\ref{fig:main_example}.
This example combines several smaller examples that were used in an exercise session to teach Excel into one spreadsheet.
Figure~\ref{fig:sol_example} shows the constraints that we expect to find.

\begin{figure}
  \label{fig:sol_example}
  {\small
    \begin{align*}
      & SERIES(\range{T_{1}}{\rangeall}{1}) \\
%
      & \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      & \ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecaggc{AVERAGE}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MAX}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MIN}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggif{SUM}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \ecaggif{MAX}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      & \range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}
  }
  \caption{Expected constraints case study}
\end{figure}
